Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2016-01-12 jss/article.tex
[skip ci] abstract and acknowledgments

%% almost as usual
\author{Paul Deveau\\Institut Curie\\PSL Research University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
        Eric Bonnet\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech}
\title{Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Paul Deveau, Eric Bonnet} %% comma-separated
\Plaintitle{Calculating Module Enrichment and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCel} %% without formatting
\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
  The abstract of the article.
}
\Keywords{keywords, comma-separated, not capitalized, \proglang{Java}}
\Plainkeywords{keywords, comma-separated, not capitalized, Java} %% without formatting
%% at least one keyword must be supplied

@@

We have recently released the Atlas of Cancer Signalling Network (ACSN), a
web-based database which describes signaling and regulatory molecular processes
that occur in a healthy mammalian cell but that are frequently deregulated
during cancerogenesis \citep{kuperstein2013navicell}.  The ACSN atlas aims to
be a comprehensive description of cancer-related mechanisms retrieved from the
most recent literature. The web interface for ACSN is using the NaviCell
technology, a software framework dedicated to web-based visualization and
navigation for biological pathway maps \citep{kuperstein2013navicell}. This
environment is providing an easy navigation of maps through the use of the
Google Maps JavaScript library, a community interface with a web blog system,
and a comprehensive module for visualization and analysis of high-throughput
data \citep{bonnet2015navicell}.

@@

\section[Examples]{Examples}

@@

>>>>

%% almost as usual
\author{Paul Deveau\\Institut Curie\\PSL Research University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
        Eric Bonnet\\Institut Curie\\PSL Research University\\INSERM U900\\Mines-ParisTech}
\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Paul Deveau, Eric Bonnet} %% comma-separated
\Plaintitle{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages ACSNMineR and RNaviCell} %% without formatting
\Shorttitle{Module Enrichment and Data Visualization with ACSNMineR and RNaviCell} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
Biological pathways or modules represent sets of interactions or functional
relationships occurring at the molecular level in living cells. A large body of
knowledge on pathways is organized in public databases such as the KEGG,
Reactome, MINT databases, but also in more specialized repositories such as the
recently released Atlas of Cancer Signaling Networks (ACSN). All these open and
popular biological databases facilitate a large spectrum of analyses, thereby
improving our understanding of cellular systems. Here we describe the
\pkg{ACSNMineR}, a novel R package for calculation of enrichment or depletion
of list of genes of interest in biological pathways. The package has built-in
support for ACSN molecular pathways, but can use any molecular pathway database
encoded as a GMT file, such as any of the gene signatures available in the
Molecular Signatures Database (MSigDB). We also present the R package
\pkg{RNaviCell}, that can be used in conjunction with \pkg{ACSNMineR} to
visualize different data types on web-based, interactive ACSN maps. We
illustrate the functionalities of the two packages with real-world biological
data taken from large-scale cancer datasets.

}
\Keywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, \proglang{R}}
\Plainkeywords{biological module, enrichment, depletion, cancer, maps, molecular networks, molecular pathways, R} %% without formatting
%% at least one keyword must be supplied

@@

We have recently released the Atlas of Cancer Signalling Networks (ACSN), a
web-based database which describes signaling and regulatory molecular processes
that occur in a healthy mammalian cell but that are frequently deregulated
during cancerogenesis \citep{kuperstein2013navicell}.  The ACSN atlas aims to
be a comprehensive description of cancer-related mechanisms retrieved from the
most recent literature. The web interface for ACSN is using the NaviCell
technology, a software framework dedicated to web-based visualization and
navigation for biological pathway maps \citep{kuperstein2013navicell}. This
environment is providing an easy navigation of maps through the use of the
Google Maps JavaScript library, a community interface with a web blog system,
and a comprehensive module for visualization and analysis of high-throughput
data \citep{bonnet2015navicell}.

@@

\section[Case studies]{Case studies}

@@

\section[Acknowledgments]{Acknowledgements}

The authors would like to thank Pierre Gestraud for his comments on early
versions of the \pkg{ACSNMineR} package and Eric Viara for guidance and
assistance on the development of the \pkg{RNaviCell} package.
